Downstream analyses of whole-genome bisulfite-sequencing data

### 2015-05-03 Rmarkdown/comethylation.Rmd
Editing
Knocking off TODOs and formatting glitches

The odds ratio is a simple summary of the dependence in a $2 \times 2$ table. An attractive feature is that it is independent of the marginal probabilities (__TODO: cite__). The odds ratio is not perfect, however. For one, the possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios. Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\theta = \log{\psi}$. The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler. I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting here that we can always convert one to the other by appropriate exponentiation or taking of logarithms.

@@

Under asymptotic environment $A2$, where the $2 \times 2$ tables are sparse, \citet{Liang:1985uw} developed three tests for the assumption of homogeneity of odds ratios and compared these with two tests for the same assumption but designed under $A1$. Unsurprisingly, \citeauthor{Liang:1985uw} found that the three conditional tests are more appropriate than either of the unconditional tests when the data are simulated under $A2$. The disadvantage of these conditional tests is the extra computation required. \citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i +} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.

- __TODO: Find the reference (I think it's Agresti) for the discussion on the utility (or lack thereof) of homogeneity tests; i.e., rejecting the null is good evidence the null is wrong but failing to reject the null is not strong evidence that it is not wrong.__

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/beta_marginals.pdf}
\caption[$Beta(0.3, 0.2)$ histogram]{Histogram of 1000 points simulated from a $Beta(0.3, 0.2)$ distribution.}
\label{fig:beta_marginals}
\end{figure}

@@

__TODO: Figure out what's going on with some of the EPISCOPE and Ziller samples that have a high autosome (chr21, I think); figure out why E18BUF is missing chrM in $NIL \geq 0$ plots.__

>>>>

The odds ratio is a simple summary of the dependence in a $2 \times 2$ table. An attractive feature is that it is independent of the marginal probabilities. The odds ratio is not perfect, however. For one, the possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios. Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\theta = \log{\psi}$. The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler. I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting here that we can always convert one to the other by appropriate exponentiation or taking of logarithms.

@@

Under asymptotic environment $A2$, where the $2 \times 2$ tables are sparse, \citet{Liang:1985uw} developed three tests for the assumption of homogeneity of odds ratios and compared these with two tests for the same assumption but designed under $A1$. Unsurprisingly, \citeauthor{Liang:1985uw} found that the three conditional tests are more appropriate than either of the unconditional tests when the data are simulated under $A2$. The disadvantage of these conditional tests is the extra computation required. \citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i +} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level. It has been noted, however, that statistical tests of homogeneity versus heterogeneity typically have low power and are only able to detect large departures form homogeneity \cite[e.g.,][]{Hauck:1989vt}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/beta_marginals.pdf}
\caption[Histogram of $Beta(0.3, 0.2)$ random variables]{Histogram of 1000 points simulated from a $Beta(0.3, 0.2)$ distribution.}
\label{fig:beta_marginals}
\end{figure}

@@

### 2015-05-03 Rmarkdown/comethylation_review.Rmd
Knocking off TODOs and formatting glitches

Specifically, \citet{Cokus:2008fc} only estimate the marginal probability that a CHH is methylated, $Pr(Z_{h, i} = 1)$, and the probability that two CHH loci on the same fragment are both methylated, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $i < i'$. From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y), x, y \in \{0, 1\}$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$. It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$. While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.

>>>>

Specifically, \citet{Cokus:2008fc} only estimate the marginal probability that a CHH is methylated, $Pr(Z_{h, i} = 1)$, and the probability that two CHH loci on the same fragment are both methylated, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $i < i'$. From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y), x, y \in \{0, 1\}$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$. It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$. While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{sec:wf_cometh} that it is more useful and simpler to estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.

### 2015-05-03 Rmarkdown/datasets.Rmd
Knocking off TODOs and formatting glitches

1. Genetic variants were called using the `bissnp_easy_usage.pl` script included with `Bis-SNP (v0.82.2)`.
2. M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set. These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e. $|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
3. CpG methylation calling was performed using `methtuple (v1.4.0)` for m-tuples $\text{m} = 1, \ldots, 8$. CpG 2-tuples were called both with and without the `--all-combinations` flag; all other tuple sizes were called without the `--all-combinations` flag.
4. Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see \ref{sec:MethylationTuples}). Specifically, a `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples, 2ac-tuples (2-tuples with `--all-combinations`), 3-tuples and 4-tuples. I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.

__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e. going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $> 275$ GB alone.

>>>>

1. Genetic variants were called using the `bissnp_easy_usage.pl` script included with `Bis-SNP (v0.82.2)`.
2. M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set. These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e. $|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
3. CpG methylation calling was performed using `methtuple (v1.4.0)` for m-tuples $\text{m} = 1, \ldots, 8$. CpG 2-tuples were called both with and without the `--all-combinations` flag; all other tuple sizes were called without the `--all-combinations` flag.
4. Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see \ref{sec:MethylationTuples}). Specifically, a `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples and 2ac-tuples (2-tuples with the  `--all-combinations` flag in `methtuple` set), 3-tuples and 4-tuples. I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.

The raw data for the _Lister_, _Seisenberger and _EPISCOPE_ datasets are all publicly available. The _EPISCOPE_ data are not yet published and I do not have permission to make these publicly available. The scripts used to prepare the results for each chapter are available from \url{https://github.com/PeteHaitch/phd_thesis_analyses}. Further details of software used are available in Appendix \ref{app:software_details}.

### 2015-05-03 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Knocking off TODOs and formatting glitches

I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple`, described in Section \ref{section:methtuple}).

@@

Now that we've found read-positions with M-bias, what can we do about it? Typically, read-positions showing evidence of M-bias are excluded when calling methylation events. In fact, a slightly cruder procedure is typically used whereby the entire ends of reads are removed. For example, suppose we have $100$ bp reads with M-bias observed at positions $1-4, 9, 94, 96-100$, then read-positions $1-9$ and $94-100$ would be ignored when methylation calling.  Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method.

>>>>

I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple`, described in Section \ref{sec:methtuple}).

@@

Now that we've found read-positions with M-bias, what can we do about it? Typically, read-positions showing evidence of M-bias are excluded when calling methylation events. In fact, a slightly cruder procedure is typically used whereby the entire ends of reads are removed. For example, suppose we have $100$ bp reads with M-bias observed at positions $1, 2, 3, 4, 9, 94, 96, 97, 98, 99, 100$, then read-positions $1-9$ and $94-100$ would be ignored when methylation calling.  Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method.

### 2015-05-03 Rmarkdown/wgbs_downstream_analyses.Rmd
Knocking off TODOs and formatting glitches

[^not_beta_values]: It is insufficient to use $\mathbf{\beta} = \frac{\mathbf{m}}{\mathbf{m} + \mathbf{u}}$ because the conversion to $\beta$-values loses information about the precision with which each methylation locus is measured (the sequencing depth, $\mathbf{d} = \mathbf{m} + \mathbf{u}$).

@@

Several software packages are now available for identifying DMCs. Most are limited to analysing two-group experiments. Rather than directly analysing the $\mathbf{m}$ and $\mathbf{u}$, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of $\mathbf{\beta}$ (Section \ref{sec:transformations}).

@@

All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\mathbf{\beta}$.

@@

Methylation data in the form of the vector of $\beta$-values, $\mathbf{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}. One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs). Partially methylated domains are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.2$ to $0.7$.

>>>>

[^not_beta_values]: It is insufficient to use $\bm{\beta} = \frac{\mathbf{m}}{\mathbf{m} + \mathbf{u}}$ because the conversion to $\beta$-values loses information about the precision with which each methylation locus is measured (the sequencing depth, $\mathbf{d} = \mathbf{m} + \mathbf{u}$).

@@

Several software packages are now available for identifying DMCs. Most are limited to analysing two-group experiments. Rather than directly analysing the $\mathbf{m}$ and $\mathbf{u}$, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of $\bm{\beta}$ (Section \ref{sec:transformations}).

@@

All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\bm{\beta}$.

@@

Methylation data in the form of the vector of $\beta$-values, $\bm{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}. One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs). Partially methylated domains are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.2$ to $0.7$.
